---
title: "PUMS Exploration"
format: 
  html:
    toc: true
    toc-depth: 3
    page-layout: full
    css: styles.css
editor: visual
---

<!-- ::: panel-tabset -->
## Variables

Exploring one-person occupancies by income

**BINCOME**: income by PSRC breakpoints

**NP**: number persons associate with housing record

**BLD**: units in structure

-   01: Mobile home or trailer

-   02: One-family house detached

-   03: One-family house attached

**HHT**: household/family type

-   4: Nonfamily household: Male householder: Living alone

-   6: Nonfamily household: Female householder: Living alone

**TEN**: tenure

-   3: Renter

-   4: Occupied without payment of rent

**BIN_AGE**: age (AGEP) grouped by standard PSRC breakpoints

**AGEP**: age of person

**SEX**: Male or Female

**VEH**: Vehicles available 

- 0: No vehicles

## Results: Overview

```{r setup}
#| echo: false
#| warning: false
#| message: false

library(psrccensus)
library(tidyverse)
library(DT)
library(psrcplot)
library(showtext)
```

```{r functions}
#| echo: false

create_dt <- function(file_name, category_var) {
  df <- data.table::fread(file.path("data", file_name))
  
  dt <- datatable(df,
          colnames = c("PUMS End Year", "Jurisdiction", category_var, "Count", "Count MOE", "Share", "Share MOE"),
           options = list(dom = 't')) |>  # default is 'lftipr', exclude f to remove search bar 
                             
  formatPercentage(c('share', 'share_moe'), 2) |> 
  formatRound(c('count', 'count_moe'), 0)
  
  return(list(df = df, dt = dt))
}

create_plot <- function(table, xvar, yvar, title, caption) {
  font_add_google("Poppins")
  showtext_auto()

  ggplot(table, aes(x = {{xvar}}, y = {{yvar}}, fill = {{xvar}})) +
    geom_col() +
    geom_text(aes(label = scales::percent(round(.data[["share"]], 2))), 
              vjust = -.5, 
              size = 25/.pt) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = psrcplot::psrc_colors$pognbgy_10, guide="none") +
    labs(x = NULL,
         y = NULL,
         title = title,
         caption = caption) +
    theme(text = element_text(size = 30, family="Poppins"),
          panel.background = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank())
}



```

```{r}
#| eval: false

x <- get_psrc_pums(span = 5,                     
                   dyear = 2023,                 
                   level = "h",                  
                   vars = c("BINCOME", "NP", "BLD", "TEN", "HHT", "BIN_AGE", "SEX", "VEH"))

# Original

# df <- psrc_pums_count(x|>
#   # filter(NP == 1 & TEN %in% c(3,4)),
#   filter(NP == 1 & TEN %in% c("Rented", "Occupied without payment of rent")),
#   stat_var = "NP",
#   group_vars = c("BINCOME")
#   )
# write.csv(df, "data/single-renter-by-income.csv")

```

```{r table}
#| echo: false

df <- data.table::fread("data/single-renter-by-income.csv")
datatable(df,
          colnames = c("PUMS End Year", "Jurisdiction", "Income", "Count", "Count MOE", "Share", "Share MOE")) |> 
  formatPercentage(c('share', 'share_moe'), 2) |> 
  formatRound(c('count', 'count_moe'), 0)
```

```{r plot df}
#| echo: false

df2 <- df |> 
  filter(BINCOME != 'Total') |> 
  mutate(BINCOME = factor(BINCOME, levels = df$BINCOME))


```

```{r plot visual, fig.width = 10, fig.height=5}
#| echo: false

font_add_google("Poppins")
showtext_auto()
                     
p <- ggplot(df2, aes(x = BINCOME, y = share, fill = BINCOME)) +
  geom_col() +
  geom_text(aes(label = scales::percent(round(share, 2))), 
            vjust = -.5, 
            size = 25/.pt) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = psrcplot::psrc_colors$pognbgy_5, guide="none") +
  labs(x = NULL,
       y = NULL,
       title = "Share of Persons Living Alone in Rented Space by Income",
       caption = "PUMS 2019-2023") +
   theme(text = element_text(size = 30, family="Poppins"),
         panel.background = element_blank(),
         axis.ticks.x = element_blank(),
         axis.ticks.y = element_blank())


p
```

## Results: Renters Earning Under $50,000

### Renters Earning by Aggregated Income
```{r table under 50k}
#| echo: false

dt_inc_agg <- create_dt("single-renter-by-income-agg.csv", category_var = "Income")

dt_inc_agg$dt
```

```{r plot under 50k, fig.width = 10, fig.height=5}
#| echo: false

dt_inc_agg2 <- dt_inc_agg$df |> 
  filter(bincome_w_50_under != 'Total') |> 
  mutate(bincome_w_50_under = factor(bincome_w_50_under, levels = dt_inc_agg$df$bincome_w_50_under))

create_plot(table = dt_inc_agg2, 
            xvar = bincome_w_50_under, 
            yvar = share, 
            title = "Share of Renters Living Alone by Income", 
            caption = "PUMS 2019-2023")
```

### Age

```{r table under 50k by age}
#| echo: false

dt_under_50k_age <- create_dt("single-renter-under-50k-by-age.csv", category_var = "Age")

dt_under_50k_age$dt

```

```{r plot under 50k by age, fig.width = 10, fig.height=5}
#| echo: false

dt_under_50k_age2 <- dt_under_50k_age$df |> 
  filter(bin_age_broad != 'Total') |> 
  mutate(bin_age_broad = factor(bin_age_broad, levels = dt_under_50k_age$df$bin_age_broad))

create_plot(table = dt_under_50k_age2, 
            xvar = bin_age_broad, 
            yvar = share, 
            title = "Share of Renters Living Alone earning Under $50,000 by Age", 
            caption = "PUMS 2019-2023")

```


### Sex

```{r table under 50k by sex}
#| echo: false

dt_under_50k_sex <- create_dt("single-renter-under-50k-by-sex.csv", category_var = "Sex")

dt_under_50k_sex$dt
```
```{r plot under 50k by sex, fig.width = 10, fig.height=5}
#| echo: false

dt_under_50k_sex2 <- dt_under_50k_sex$df |> 
  filter(SEX != 'Total') |> 
  mutate(SEX = factor(SEX, levels = dt_under_50k_sex$df$SEX))

create_plot(table = dt_under_50k_sex2, 
            xvar = SEX, 
            yvar = share, 
            title = "Share of Renters Living Alone earning Under $50,000 by Sex", 
            caption = "PUMS 2019-2023")

```


### Number of Vehicles

```{r table under 50k by vehicles}
#| echo: false

dt_under_50k_veh <- create_dt("single-renter-under-50k-by-veh.csv", category_var = "Number of Vehicles")

dt_under_50k_veh$dt
```
```{r plot under 50k by vehicles, fig.width = 10, fig.height=5}
#| echo: false

dt_under_50k_veh2 <- dt_under_50k_veh$df |> 
  filter(vehicles != 'Total') |> 
  mutate(vehicles = factor(vehicles, levels = dt_under_50k_veh$df$vehicles))

create_plot(table = dt_under_50k_veh2, 
            xvar = vehicles, 
            yvar = share, 
            title = "Share of Renters Living Alone earning Under $50,000 by Number of Vehicles", 
            caption = "PUMS 2019-2023")
```

<!-- ::: -->
