---
title: "PUMS Exploration"
author: "Christy"
format: 
  html:
    page-layout: full
editor: visual
---

## PUMS

Exploring one-person households by income

## Variables

**BINCOME**: income by PSRC breakpoints

**NP**: number persons associate with housing record

**BLD**: units in structure

-   01: Mobile home or trailer

-   02: One-family house detached

-   03: One-family house attached

**HHT**: household/family type

-   4: Nonfamily household: Male householder: Living alone

-   6: Nonfamily household: Female householder: Living alone

**TEN**: tenure

-    3: Renter

-   4: Occupied without payment of rent

```{r setup}
#| echo: false
#| warning: false

library(psrccensus)
library(tidyverse)
library(DT)
library(psrcplot)
library(showtext)
```

```{r}
#| eval: false

x <- get_psrc_pums(span = 5,                     
                   dyear = 2023,                 
                   level = "h",                  
                   vars = c("BINCOME", "NP", "BLD", "TEN", "HHT"))

df <- psrc_pums_count(x|> 
  filter(NP == 1 & TEN %in% c(3,4)), 
  stat_var = "NP", 
  group_vars = c("BINCOME")
  )

# write.csv(df, "data/single-renter-by-income.csv")

```

```{r table}
df <- data.table::fread("data/single-renter-by-income.csv")
datatable(df,
          colnames = c("PUMS End Year", "Jurisdiction", "Income", "Count", "Count MOE", "Share", "Share MOE")) |> 
  formatPercentage(c('share', 'share_moe'), 2) |> 
  formatRound(c('count', 'count_moe'), 0)
```

```{r plot df}

df2 <- df |> 
  filter(BINCOME != 'Total')
```

```{r plot visual, fig.width = 10, fig.height=5}

font_add_google("Poppins")
showtext_auto()

p <- ggplot(df2, aes(x = BINCOME, y = share, fill = BINCOME)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = psrcplot::psrc_colors$pognbgy_5, guide="none") +
  labs(x = NULL,
       y = NULL,
       title = "Share of Persons Living Alone in Rented Space by Income",
       caption = "PUMS 2019-2023") +
   theme(text = element_text(size=36, family="Poppins"),
         panel.background = element_blank(),
         axis.ticks.x = element_blank(),
         axis.ticks.y = element_blank())


p
```

